Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-09-04 old-version/paper.tex
added the old (long) version of the paper

% Set up for ASA data science journal
\articletype{Article Type}
\received{26 April 2016}
\revised{6 June 2016}
\accepted{6 June 2016}
\raggedbottom

\begin{document}

@@

<<setup, echo = FALSE, include = FALSE>>=

# Specify global options
knitr::opts_chunk$set(echo = FALSE,
                      fig.keep = 'high',
                      message = FALSE)

# Load packages
library(caret)
library(cowplot)
library(dplyr)
library(forcats)
library(ggforce)
library(ggplot2)
library(ggpcp)
library(glmnet)
library(gower)
library(graphics)
library(gretchenalbrecht)
library(latex2exp)
library(lime) #devtools::install_github("goodekat/lime")
library(limeaid) #devtools::install_github("goodekat/limeaid")
library(purrr)
library(randomForest)
library(stringr)
library(tidyr)

@

@@

<<concept-data>>=
# Simulate example data
set.seed(20190624)
lime_data <-
  data.frame(
    feature1 = sort(runif(250, 0, 1)),
    feature2 = sample(x = 1:250, size = 250, replace = FALSE)) %>%
  mutate(
    feature1_stnd = (feature1 - mean(feature1)) / sd(feature1),
    feature2_stnd = (feature2 - mean(feature2)) / sd(feature2),
    prediction = if_else(feature1 >= 0 & feature1 < 0.1,
                         (0.3 * feature1) + rnorm(n(), 0, 0.01),
                 if_else(feature1 >= 0.1 & feature1 < 0.3,
                         rbeta(n(), 1, 0.5),
                 if_else(feature1 >= 0.3 & feature1 < 0.5,
                         sin(pi* feature1) + rnorm(n(), 0, 0.5),
                 if_else(feature1 >= 0.5 & feature1 < 0.8,
                         -(sin(pi* feature1) + rnorm(n(), 0, 0.1)) + 1,
                 if_else(feature1 >= 0.8 & feature1 < 0.9,
                         0.5 + runif(n(), -0.5, 0.5),
                         0.5 + rnorm(n(), 0, 0.3))))))) %>%
  bind_rows(
    data.frame(feature1 = rep(1, 30) + rnorm(5, 0, 0.05),
           feature2 = rep(245, 30) + rnorm(5, 0, 1),
           prediction = rep(0, 30) + rnorm(5, 0, 0.05)) %>%
  mutate(feature1_stnd = (feature1 - mean(feature1)) / sd(feature1),
         feature2_stnd = (feature2 - mean(feature2)) / sd(feature2)))

# Specify a prediction of interest
prediction_of_interest <-
  data.frame(feature1 = 0.07, feature2 = 200) %>%
  mutate(feature1_stnd = (feature1 - mean(lime_data$feature1)) /
           sd(lime_data$feature1),
         feature2_stnd = (feature2 - mean(lime_data$feature2)) /
           sd(lime_data$feature2),
         prediction = 0.05,
         color = factor("Prediction \nof Interest"))

# Specify the gower exponents
good_gower_power <- 50
bad_gower_power <- 1

# Compute the good distances between the prediction of interest
# and all other observations
lime_data$distance_good <-
  (1 - gower_dist(x = prediction_of_interest %>%
                    select(feature1_stnd, feature2_stnd),
                  y = lime_data %>%
                    select(feature1_stnd,
                           feature2_stnd)))^good_gower_power

# Compute the bad distances between the prediction of interest
# and all other observations
lime_data$distance_bad <-
  (1 - gower_dist(x = prediction_of_interest %>%
                    select(feature1_stnd, feature2_stnd),
                  y = lime_data %>%
                    select(feature1_stnd,
                           feature2_stnd)))^bad_gower_power

# Prepare the data for plotting
lime_data_gathered <- lime_data %>%
  gather(feature, feature_stnd_value,
         feature1_stnd:feature2_stnd) %>%
  gather(distance, distance_value, distance_good:distance_bad) %>%
  mutate(distance = fct_recode(distance,
                               "good" = "distance_good",
                               "bad" = "distance_bad"),
         feature = fct_recode(feature,
                              "Feature 1" = "feature1_stnd",
                              "Feature 2" = "feature2_stnd"))

# Prepare the prediction of interest data for plotting
prediction_of_interest_gathered <- prediction_of_interest %>%
  select(-feature1, -feature2) %>%
  gather(feature, feature_value, feature1_stnd, feature2_stnd) %>%
  mutate(feature = fct_recode(feature,
                              "Feature 1" = "feature1_stnd",
                              "Feature 2" = "feature2_stnd"))

@

<<concept-explainers>>=

# Fit the good interpretable explainer model
explainer_good <-
  glmnet(x = lime_data %>% select(feature1_stnd, feature2_stnd)
         %>% as.matrix(),
         y = lime_data$prediction,
         alpha = 0,
         lambda = 1,
         weights = lime_data$distance_good)

# Fit the bad interpretable explainer model
explainer_bad <-
  glmnet(x = lime_data %>% select(feature1_stnd, feature2_stnd)
         %>% as.matrix(),
         y = lime_data$prediction,
         alpha = 0,
         lambda = 1,
         weights = lime_data$distance_bad)

# Join the coefficients from the explainer model into a dataframe
coefs_data <- data.frame(case = c("good", "bad"),
                         b0 = c(coef(explainer_good)[1],
                                coef(explainer_bad)[1]),
                         b1 = c(coef(explainer_good)[2],
                                coef(explainer_bad)[2]),
                         b2 = c(coef(explainer_good)[3],
                                coef(explainer_bad)[3])) %>%
  mutate(feature1 = b0 + b2*prediction_of_interest$feature2_stnd,
         feature2 = b0 + b1*prediction_of_interest$feature1_stnd) %>%
  gather(key = feature, value = int, feature1:feature2) %>%
  mutate(slope = c(coef(explainer_good)[2],
                   coef(explainer_good)[3],
                   coef(explainer_bad)[2],
                   coef(explainer_bad)[3]),
         feature = fct_recode(feature,
                              "Feature 1" = "feature1",
                              "Feature 2" = "feature2"))

@

\begin{figure*}[!htbp]
<<concept-plot-good, out.width = '\\textwidth', fig.width = 7.95, fig.height = 3.75, warning = FALSE>>=

# Plot of good explainer model
ggplot() +
  facet_grid(. ~ feature, switch = "x") +
  geom_point(data = lime_data_gathered,
             mapping = aes(x = feature_stnd_value,
                           y = prediction,
                           size = distance_value,
                           alpha = distance_value,
                           color = distance)) +
  geom_abline(data = coefs_data %>% filter(case == "good"),
              mapping = aes(intercept = int, slope = slope),
              size = 1) +
  scale_alpha(range = c(0.3, 1)) +
  scale_color_manual(values = c(NA, "grey30"), guide = FALSE) +
  geom_point(data = prediction_of_interest_gathered,
             mapping = aes(x = feature_value,
                           y = prediction,
                           fill = color),
             color = "black",
             size = 5,
             shape = 23,
             alpha = 0.75) +
  scale_fill_manual(values = "#FAAA72") +
  geom_text(
    data = data.frame(feature_stnd_value = 1.2,
                      prediction = 1.6,
                      feature = c("Feature 1", "Feature 2"),
                      slope = c(paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "good",
                                               feature == "Feature 1") %>%
                                        pull(slope) %>%
                                        round(3)),
                                paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "good",
                                               feature == "Feature 2") %>%
                                        pull(slope) %>%
                                        round(3)))),
    mapping = aes(x = feature_stnd_value,
                  y = prediction,
                  label = slope),
    family = "Times",
    size = 4) +
  labs(x = "",
       y = "Black-Box Prediction",
       title = "Conceptual Depiction of a Faithful Local Explainer Model",
       subtitle = paste("Gower Distance Metric Exponent:", good_gower_power),
       fill = "",
       alpha = "Weight",
       size = "Weight") +
  theme_linedraw(base_family = "Times", base_size = 10) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.placement = "outside",
        strip.background = element_rect(color = "white", fill = "white"),
        strip.text = element_text(color = 'black', size = 10)) +
  guides(fill = guide_legend(order = 1),
         size = guide_legend(order = 2),
         alpha = guide_legend(order = 2))

@
\caption{A conceptual depiction of a faithful LIME explainer model in the immediate neighborhood of a prediction of interest. The predictions from a hypothetical black-box model are plotted against the standardized values of the two predictor variables used to fit the hypothetical model. The diamond shaped points represent the location of a prediction of interest. The size and opacity of the circular points indicate the weight assigned based on the distance to the point of interest computed using the inverse of the Gower distance metric raised to the power of \Sexpr{good_gower_power}. The black lines are a weighted ridge regression model used as an explainer model that reasonably captures the relationship between the black-box predictions and the features in a local region around the point of interest. That is, it is faithful to the complex model and produces a reasonable explanation that Feature 1 plays a more important role in the prediction of interest than Feature 2 since the magnitude of the slope associated with Feature 1 is larger.}
\label{fig:concept-plot-good}

@@

<<concept-plot-bad, out.width = '\\textwidth', fig.width = 7.95, fig.height = 3.75, warning = FALSE>>=

# Plot of good explainer model
ggplot() +
  facet_grid(. ~ feature, switch = "x") +
  geom_point(data = lime_data_gathered,
             mapping = aes(x = feature_stnd_value,
                           y = prediction,
                           size = distance_value,
                           alpha = distance_value,
                           color = distance)) +
  geom_abline(data = coefs_data %>% filter(case == "bad"),
              mapping = aes(intercept = int, slope = slope),
              size = 1) +
  scale_alpha(range = c(0.3, 1)) +
  scale_color_manual(values = c("grey30", NA), guide = FALSE) +
  geom_point(data = prediction_of_interest_gathered,
             mapping = aes(x = feature_value,
                           y = prediction,
                           fill = color),
             color = "black",
             size = 5,
             shape = 23,
             alpha = 0.75) +
  scale_fill_manual(values = "#FAAA72") +
  geom_text(
    data = data.frame(feature_stnd_value = 1.2,
                      prediction = 1.6,
                      feature = c("Feature 1", "Feature 2"),
                      slope = c(paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "bad",
                                               feature == "Feature 1") %>%
                                        pull(slope) %>%
                                        round(3)),
                                paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "bad",
                                               feature == "Feature 2") %>%
                                        pull(slope) %>%
                                        round(3)))),
    mapping = aes(x = feature_stnd_value,
                  y = prediction,
                  label = slope),
    family = "Times",
    size = 4) +
  labs(x = "",
       y = "Black-Box Prediction",
       title = "Conceptual Depiction of an Unfaithful Local Explainer Model",
       subtitle = paste("Gower Distance Metric Exponent:", bad_gower_power),
       fill = "",
       alpha = "Weight",
       size = "Weight") +
  theme_linedraw(base_family = "Times", base_size = 10) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.placement = "outside",
        strip.background = element_rect(color = "white", fill = "white"),
        strip.text = element_text(color = 'black', size = 10)) +
  guides(fill = guide_legend(order = 1),
         size = guide_legend(order = 2),
         alpha = guide_legend(order = 2))

@
\caption{A conceptual depiction of an unfaithful local explainer model in the immediate neighborhood of a prediction of interest. A ridge regression model is fit to the same black-box model predictions and predictor variables as in \autoref{fig:concept-plot-good}, but the weights are computed using a Gower exponent of 1. The explainer model provides a global approximation of the complex model in regards to the prediction of interest.}
\label{fig:concept-plot-bad}
\end{figure*}

@@

\autoref{fig:concept-plot-good} provides a visualization of this conceptual understanding of LIME. The two plots show the predictions from a hypothetical black-box model plotted against the two predictor variables used to fit the hypothetical model. The diamond shaped points represent a prediction  of interest. A Gower distance metric \citep{gower:1971} is used to define locality: the size of the points represent the inverse of the Gower distance as a reflection of the proximity to the prediction of interest. Here, an exponent of \Sexpr{good_gower_power} is used to emphasize interest in a very local region around the point of interest. A ridge regression model weighted by the proximity values is used as the explainer model. The black-box predictions are used as the response variable, standardized versions of Features 1 and 2 are included as predictor variables. Standardized features allow direct comparisons of the model coefficients. The explainer model is depicted by the black lines in the figure.

The plot on the left of \autoref{fig:concept-plot-good} shows the relationship between the black-box predictions and Feature 1.  Here, the explainer model is plotted with Feature 2 fixed at the observed value of Feature 2 for the prediction of interest. The explainer model captures the relationship in the immediate neighborhood around the prediction of interest  with a slope of \Sexpr{coefs_data %>% filter(case == "good", feature == "Feature 1") %>% pull(slope) %>% round(3)}. The plot on the right shows that there is no global or local relationship between the black-box predictions and Feature 2. Here, the explainer model is plotted with Feature 1 set to be the observed value of Feature 1 for the prediction of interest, and it has a small slope of \Sexpr{coefs_data %>% filter(case == "good", feature == "Feature 2") %>% pull(slope) %>% round(3)}. The magnitude of the slope associated with Feature 1 is larger than the slope of Feature 2, resulting LIME to report that Feature 1 plays a more important role in the prediction made by the black-box model.

@@

The choice of exponent in the scenario depicted in \autoref{fig:concept-plot-good} is crucial. High values ensure an up-weight of observations closer to the prediction of interest. Now consider the example depicted in \autoref{fig:concept-plot-bad}. The plots show the same data as \autoref{fig:concept-plot-good}, but the Gower distance metric was raised to the default power used by the \emph{lime} R package of \Sexpr{bad_gower_power}. This causes the observations that are further away from the prediction of interest to be given larger weights than before. In \autoref{fig:concept-plot-good}, the explainer model captures the relationship between the black-box predictions in the immediate neighborhood of the prediction of interest. This cannot be said of the explainer model in \autoref{fig:concept-plot-bad}. In addition, the magnitude of the slope (\Sexpr{coefs_data %>% filter(case == "bad", feature == "Feature 2") %>% pull(slope) %>% round(3) %>% abs()}) associated with Feature 2 is larger than that of Feature 1 (\Sexpr{coefs_data %>% filter(case == "bad", feature == "Feature 1") %>% pull(slope) %>% round(3) %>% abs()}). Thus, this explainer model actually provides a misleading explanation that Feature 2 plays a more important role in predicting the observation of interest.

@@

LIME was initially implemented in a Python package\footnote{https://github.com/marcotcr/lime} by the original authors. The adaption to an R package is by Thomas Lin-Pedersen\footnote{https://github.com/thomasp85/lime}. In this paper, we use a forked version\footnote{https://github.com/goodekat/lime} of the development version of the R package \emph{lime} (\Sexpr{packageVersion("lime")}) \citep{pedersen:2020}. The forked version of \emph{lime} allows us to export internal values of the procedure not already exported by \emph{lime}.

@@

<<sine-data, echo = FALSE, include = FALSE>>=

# Functions for rotating the data
rot_x <- function(x, y, theta) (x * cos(theta)) - (y * sin(theta))
rot_y <- function(x, y, theta) (x * sin(theta)) + (y * cos(theta))

# Generate the data
theta = -0.9
min = -10
max = 10
set.seed(20190913)
sine_data <- data.frame(x1 = runif(600, min, max),
                        x2 = sort(runif(600, min, max))) %>%
  mutate(x1new = rot_x(x1, x2, theta),
         x2new = rot_y(x1, x2, theta),
         y = ifelse(x2new > 5 * sin(x1new), "blue", "red")) %>%
  slice(sample(1:n())) %>%
  mutate(x3 = rnorm(600),
         case = 1:600) %>%
  select(case, everything())

# Separte the data into training and testing parts
set.seed(20191003)
rs <- sample(1:600, 500, replace = FALSE)
sine_data_train <- sine_data[rs,]
sine_data_test <- sine_data[-rs,]

# Fit a random forest
set.seed(20191003)
rfsine <- randomForest(x = sine_data_train %>% select(x1, x2, x3),
                       y = factor(sine_data_train$y))

# Obtain predictions on the training and testing data
sine_data_train$rfpred <- predict(rfsine)
sine_data_train <- cbind(sine_data_train, predict(rfsine, type = "prob"))
sine_data_train <- sine_data_train %>% 
  rename(rfprob_blue = blue, rfprob_red = red)
sine_data_test$rfpred <- predict(rfsine, sine_data_test %>% select(x1, x2, x3))
sine_data_test <- cbind(sine_data_test,
                        predict(rfsine, sine_data_test
                                %>% select(x1, x2, x3),
                                type = "prob"))
sine_data_test <- sine_data_test %>% 
  rename(rfprob_blue = blue, rfprob_red = red)

# Extract the prediction of interest from the sine test data
sine_poi <- sine_data_test %>% filter(y != rfpred, x1 > 0, x1 < 5, x2 > 5)

@

\begin{figure*}[!t]
\centering
<<sine-plot-data, out.width = '\\textwidth', fig.width = 8, fig.height = 3.1>>=

# Create points represting the rotated since curve
sinefun_data <- data.frame(xnew = seq(min(sine_data$x1new),
                                      max(sine_data$x1new),
                                      by = 0.01)) %>%
  mutate(ynew = 5 * sin(xnew)) %>%
  mutate(x = rot_x(xnew, ynew, -theta),
         y = rot_y(xnew, ynew, -theta)) %>%
  filter(y >= -10, y <= 10)

# Specify colors for predictions
sinecolor_red = "firebrick"
sinecolor_blue = "steelblue"

# Plot the training data observed classes
sine_plot_obs <- ggplot(sine_data_train, aes(x = x1, y = x2, color = y)) +
  geom_point(alpha = 0.8) +
  geom_path(data = sinefun_data, aes(x = x, y = y),
            color = "black") +
  scale_colour_manual(values = c(sinecolor_blue, sinecolor_red)) +
  theme_bw(base_family = "Times", base_size = 10) +
  labs(x = TeX("$x_1$"),
       y = TeX("$x_2$"),
       color = TeX("y"),
       title = "Training Data")

# Plot the testing data rf predictions
sine_plot_pred <- ggplot() +
  geom_point(data = sine_poi %>%
               mutate(shape = "Prediction \nof Interest"),
             mapping = aes(x = x1,
                           y = x2,
                           #fill = rfprob_blue,
                           shape = shape),
             size = 5,
             alpha = 0.8,
             color = "black") +
  geom_point(data = sine_data_test %>% filter(y != rfpred),
             mapping = aes(x = x1, y = x2),
             shape = 1,
             size = 3,
             color = "black") +
  geom_point(data = sine_data_test,
             mapping =  aes(x = x1,
                            y = x2,
                            color = rfprob_blue,
                            fill = rfprob_blue),
             shape = 21,
             alpha = 0.8) +
  geom_path(data = sinefun_data, aes(x = x, y = y),
            color = "black") +
  scale_color_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5) +
  scale_fill_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5) +
  scale_shape_manual(values = 23) +
  theme_bw(base_family = "Times", base_size = 10) +
  labs(x = TeX("$x_1$"),
       y = TeX("$x_2$"),
       color = "Random \nForest \nProbability \nfor 'blue'",
       fill = "Random \nForest \nProbability \nfor 'blue'",
       title = "Testing Data",
       shape = "") +
  guides(shape = guide_legend(order = 1))

# Join the plots
plot_grid(sine_plot_obs, sine_plot_pred,
          nrow = 1,
          rel_widths = c(0.4825, 0.5175))

@
\caption{Plots of $x_2$ versus $x_1$ from the training (left) and testing (right) set of the \data \ introduced in Section \ref{procedure}. The true classification boundary is shown as the solid black line in both plots. The color of the training data points represents the value of the observed response variable ($y$). The color of the testing data points represents the random forest probability that an observation belongs to the category of blue ($\hat{y}$). The \Sexpr{dim(sine_data_test %>% filter(y != rfpred))[1]} cases that are misclassified by the random forest are identified by black circles. The prediction of interest to explain is indicated by a diamond.}
\label{fig:sine-plot-data}
\end{figure*}

@@

To demonstrate the visualizations, we generate an example dataset that will be referred to as the \data. The \data \ contains \Sexpr{dim(sine_data)[1]} observations with three features and one response variable. The features, $x_1$, $x_2$, and $x_3$, are randomly sampled from Unif(\Sexpr{min}, \Sexpr{max}), Unif(\Sexpr{min}, \Sexpr{max}), and $\mbox{N}(0,1)$ distributions, respectively. A binary response variable $y$ is  created using a rotated sine curve. In particular, let $x'_1=x_1\cos(\theta)-x_2\sin(\theta)$ and $x'_2=x_1\sin(\theta)+x_2\cos(\theta)$ where $\theta=-0.9$. Then $y$ is defined as
\begin{eqnarray}\label{eq:data}
  y=\begin{cases}
  \mbox{blue} & \mbox{ if } x'_2 > \sin\left(x'_1\right) \\
  \mbox{red} & \mbox{ if } x'_2 \le \sin\left(x'_1\right) \ . %\nolabel
  \end{cases}
\end{eqnarray}
Note that due to the creation of $y$ in this manner, $y$ is dependent on $x_1$ and $x_2$ and independent of $x_3$.

The dataset is divided into a training set of \Sexpr{dim(sine_data_train)[1]} observations and a testing set of \Sexpr{dim(sine_data_test)[1]} observations. A random forest model is fit using the R package \emph{randomForest} (version \Sexpr{packageVersion("randomForest")}) \citep{liaw:2002} with the default settings. The model is applied to the testing set to obtain predictions. \autoref{fig:sine-plot-data} shows scatterplots of $x_2$ versus $x_1$ from the training data (left) and the testing data (right). Both plots include the true classification boundary of the rotated sine function plotted as the solid black line. The training data are colored by the observed response variable ($y$), and the testing data are colored by the prediction probabilities from the random forest model ($\hat{y}$).

The random forest model misclassifies \Sexpr{dim(sine_data_test %>% filter(y != rfpred))[1]} points on the classification boundary. These are identified by circles in \autoref{fig:sine-plot-data}. For the visualizations of the steps of the LIME algorithm, we focus on one of these points: the observation with $(x_1, x_2, x_3)$ coordinates of (\Sexpr{round(sine_poi$x1, 2)}, \Sexpr{round(sine_poi$x2, 2)}, \Sexpr{round(sine_poi$x3, 2)}) and indicated in \autoref{fig:sine-plot-data} by a diamond.

<<sine-lime>>=

# Apply LIME with various tuning parameters to the sine data
if (!file.exists("../data/sine_lime_explain.rds")) {

    # Apply lime with various input options
    sine_lime_explain <- apply_lime(
      train = sine_data_train %>% select(x1, x2, x3),
      test = sine_data_test %>% select(x1, x2, x3),
      model = rfsine,
      label = "blue",
      n_features = 2,
      sim_method = c('quantile_bins', "kernel_density"),
      nbins = 2:6,
      feature_select = "auto",
      dist_fun = "gower",
      kernel_width = NULL,
      gower_pow = 1,
      return_perms = TRUE,
      all_fs = TRUE,
      seed = 20190914)

    saveRDS(sine_lime_explain, "../data/sine_lime_explain.rds")

  } else {

    sine_lime_explain <- readRDS("../data/sine_lime_explain.rds")

  }

@

<<sine-lime-default>>=

# Extract the explanations from the default lime implementation
sine_lime_default <- sine_lime_explain$explain %>% filter(nbins == 4)

# Extract the explanations from the default lime implementation
# for the prediction of interest only
sine_poi_lime_default <- sine_lime_default %>%
  filter(case == sine_poi %>% pull(case)) %>%
  mutate(case = c("Prediction of Interest", "Prediction of Interest"))

@

@@

\autoref{fig:sine-plot-poi} shows the visualization of the explanation produced by \emph{lime} for the prediction of interest. The figure shows that the random forest returned a probability of \Sexpr{sine_poi_lime_default$label_prob[1]} that the observation of interest belongs to category blue {(denoted as the ``label" in the figure). The colored bars indicate the features chosen by LIME and whether they support or contradict a random forest classification of blue. Note that the LIME explanation provides its explanations in form of intervals surrounding the observation of interest. We will discuss this aspect in more detail in Section \ref{step1}.

The explanation for the prediction of interest depicted in Figure \ref{fig:sine-plot-poi} can be interpreted as follows. A random forest classification of blue is supported by the prediction of interest having a value of $x_2$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[1]], pattern = " ")[[1]][1]}, but the prediction of interest having a value of $x_1$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][1]} and less than or equal to \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][5]} provides support against a classification of blue. Since the weight associated with $x_2$ has a larger magnitude than the weight associated with $x_1$, LIME explains that the $x_1$ and $x_2$ locations of the prediction of interest provide more support for a random forest classification of blue over red. This explanation agrees with the random forest probability of \Sexpr{sine_poi_lime_default$label_prob[1]} for blue, even though both classify the observation incorrectly.

\autoref{fig:sine-plot-poi} also includes an "explanation fit" value. This value corresponds to the deviance ratio from the R package \emph{glmnet} \citep{simon:2011} for a ridge regression model, or in other words, this is the $R^2$ value associated with the explainer model. In this case, it is \Sexpr{round(sine_poi_lime_default$model_r2[1], 2)} suggesting that the explainer model is not a good linear fit. However, it is commonly accepted that $R^2$ has limitations for accessing the quality of fit of a model \citep{sapra:2014}, and it should not be the only metric used for assessment of a model. Based on this metric or this plot, it is not yet possible to make an informed assessment of the trustworthiness of the explanation.

@@

\begin{figure*}[!htp]
\centering
<<sine-plot-step2ab, out.width = '\\textwidth', fig.width = 8, fig.height = 3.25>>=

>>>>

% Set up for ASA data science journal
\articletype{Article Type}
\received{26 April 2016}
\revised{6 June 2016}
\accepted{6 June 2016}
\raggedbottom
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

@@

\begin{figure*}[!htbp]
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{figure/concept-plot-good-1}

\end{knitrout}
\caption{A conceptual depiction of a faithful LIME explainer model in the immediate neighborhood of a prediction of interest. The predictions from a hypothetical black-box model are plotted against the standardized values of the two predictor variables used to fit the hypothetical model. The diamond shaped points represent the location of a prediction of interest. The size and opacity of the circular points indicate the weight assigned based on the distance to the point of interest computed using the inverse of the Gower distance metric raised to the power of 50. The black lines are a weighted ridge regression model used as an explainer model that reasonably captures the relationship between the black-box predictions and the features in a local region around the point of interest. That is, it is faithful to the complex model and produces a reasonable explanation that Feature 1 plays a more important role in the prediction of interest than Feature 2 since the magnitude of the slope associated with Feature 1 is larger.}
\label{fig:concept-plot-good}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{figure/concept-plot-bad-1}

\end{knitrout}
\caption{A conceptual depiction of an unfaithful local explainer model in the immediate neighborhood of a prediction of interest. A ridge regression model is fit to the same black-box model predictions and predictor variables as in \autoref{fig:concept-plot-good}, but the weights are computed using a Gower exponent of 1. The explainer model provides a global approximation of the complex model in regards to the prediction of interest.}
\label{fig:concept-plot-bad}
\end{figure*}

@@

\autoref{fig:concept-plot-good} provides a visualization of this conceptual understanding of LIME. The two plots show the predictions from a hypothetical black-box model plotted against the two predictor variables used to fit the hypothetical model. The diamond shaped points represent a prediction  of interest. A Gower distance metric \citep{gower:1971} is used to define locality: the size of the points represent the inverse of the Gower distance as a reflection of the proximity to the prediction of interest. Here, an exponent of 50 is used to emphasize interest in a very local region around the point of interest. A ridge regression model weighted by the proximity values is used as the explainer model. The black-box predictions are used as the response variable, standardized versions of Features 1 and 2 are included as predictor variables. Standardized features allow direct comparisons of the model coefficients. The explainer model is depicted by the black lines in the figure.

The plot on the left of \autoref{fig:concept-plot-good} shows the relationship between the black-box predictions and Feature 1.  Here, the explainer model is plotted with Feature 2 fixed at the observed value of Feature 2 for the prediction of interest. The explainer model captures the relationship in the immediate neighborhood around the prediction of interest  with a slope of 0.068. The plot on the right shows that there is no global or local relationship between the black-box predictions and Feature 2. Here, the explainer model is plotted with Feature 1 set to be the observed value of Feature 1 for the prediction of interest, and it has a small slope of -0.001. The magnitude of the slope associated with Feature 1 is larger than the slope of Feature 2, resulting LIME to report that Feature 1 plays a more important role in the prediction made by the black-box model.

@@

The choice of exponent in the scenario depicted in \autoref{fig:concept-plot-good} is crucial. High values ensure an up-weight of observations closer to the prediction of interest. Now consider the example depicted in \autoref{fig:concept-plot-bad}. The plots show the same data as \autoref{fig:concept-plot-good}, but the Gower distance metric was raised to the default power used by the \emph{lime} R package of 1. This causes the observations that are further away from the prediction of interest to be given larger weights than before. In \autoref{fig:concept-plot-good}, the explainer model captures the relationship between the black-box predictions in the immediate neighborhood of the prediction of interest. This cannot be said of the explainer model in \autoref{fig:concept-plot-bad}. In addition, the magnitude of the slope (0.011) associated with Feature 2 is larger than that of Feature 1 (0.005). Thus, this explainer model actually provides a misleading explanation that Feature 2 plays a more important role in predicting the observation of interest.

@@

LIME was initially implemented in a Python package\footnote{https://github.com/marcotcr/lime} by the original authors. The adaption to an R package is by Thomas Lin-Pedersen\footnote{https://github.com/thomasp85/lime}. In this paper, we use a forked version\footnote{https://github.com/goodekat/lime} of the development version of the R package \emph{lime} (0.5.1) \citep{pedersen:2020}. The forked version of \emph{lime} allows us to export internal values of the procedure not already exported by \emph{lime}.

@@

\begin{figure*}[!t]
\centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{figure/sine-plot-data-1}

\end{knitrout}
\caption{Plots of $x_2$ versus $x_1$ from the training (left) and testing (right) set of the \data \ introduced in Section \ref{procedure}. The true classification boundary is shown as the solid black line in both plots. The color of the training data points represents the value of the observed response variable ($y$). The color of the testing data points represents the random forest probability that an observation belongs to the category of blue ($\hat{y}$). The 18 cases that are misclassified by the random forest are identified by black circles. The prediction of interest to explain is indicated by a diamond.}
\label{fig:sine-plot-data}
\end{figure*}

@@

To demonstrate the visualizations, we generate an example dataset that will be referred to as the \data. The \data \ contains 600 observations with three features and one response variable. The features, $x_1$, $x_2$, and $x_3$, are randomly sampled from Unif(-10, 10), Unif(-10, 10), and $\mbox{N}(0,1)$ distributions, respectively. A binary response variable $y$ is  created using a rotated sine curve. In particular, let $x'_1=x_1\cos(\theta)-x_2\sin(\theta)$ and $x'_2=x_1\sin(\theta)+x_2\cos(\theta)$ where $\theta=-0.9$. Then $y$ is defined as
\begin{eqnarray}\label{eq:data}
  y=\begin{cases}
  \mbox{blue} & \mbox{ if } x'_2 > \sin\left(x'_1\right) \\
  \mbox{red} & \mbox{ if } x'_2 \le \sin\left(x'_1\right) \ . %\nolabel
  \end{cases}
\end{eqnarray}
Note that due to the creation of $y$ in this manner, $y$ is dependent on $x_1$ and $x_2$ and independent of $x_3$.

The dataset is divided into a training set of 500 observations and a testing set of 100 observations. A random forest model is fit using the R package \emph{randomForest} (version 4.6.14) \citep{liaw:2002} with the default settings. The model is applied to the testing set to obtain predictions. \autoref{fig:sine-plot-data} shows scatterplots of $x_2$ versus $x_1$ from the training data (left) and the testing data (right). Both plots include the true classification boundary of the rotated sine function plotted as the solid black line. The training data are colored by the observed response variable ($y$), and the testing data are colored by the prediction probabilities from the random forest model ($\hat{y}$).

The random forest model misclassifies 18 points on the classification boundary. These are identified by circles in \autoref{fig:sine-plot-data}. For the visualizations of the steps of the LIME algorithm, we focus on one of these points: the observation with $(x_1, x_2, x_3)$ coordinates of (1.23, 8.47, -0.99) and indicated in \autoref{fig:sine-plot-data} by a diamond.

@@

\autoref{fig:sine-plot-poi} shows the visualization of the explanation produced by \emph{lime} for the prediction of interest. The figure shows that the random forest returned a probability of 0.744 that the observation of interest belongs to category blue {(denoted as the ``label" in the figure). The colored bars indicate the features chosen by LIME and whether they support or contradict a random forest classification of blue. Note that the LIME explanation provides its explanations in form of intervals surrounding the observation of interest. We will discuss this aspect in more detail in Section \ref{step1}.

The explanation for the prediction of interest depicted in Figure \ref{fig:sine-plot-poi} can be interpreted as follows. A random forest classification of blue is supported by the prediction of interest having a value of $x_2$ that is greater than 4.834, but the prediction of interest having a value of $x_1$ that is greater than -0.302 and less than or equal to 4.844 provides support against a classification of blue. Since the weight associated with $x_2$ has a larger magnitude than the weight associated with $x_1$, LIME explains that the $x_1$ and $x_2$ locations of the prediction of interest provide more support for a random forest classification of blue over red. This explanation agrees with the random forest probability of 0.744 for blue, even though both classify the observation incorrectly.

\autoref{fig:sine-plot-poi} also includes an "explanation fit" value. This value corresponds to the deviance ratio from the R package \emph{glmnet} \citep{simon:2011} for a ridge regression model, or in other words, this is the $R^2$ value associated with the explainer model. In this case, it is 0.21 suggesting that the explainer model is not a good linear fit. However, it is commonly accepted that $R^2$ has limitations for accessing the quality of fit of a model \citep{sapra:2014}, and it should not be the only metric used for assessment of a model. Based on this metric or this plot, it is not yet possible to make an informed assessment of the trustworthiness of the explanation.

@@

\begin{figure*}[!htp]
\centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\textwidth]{figure/sine-plot-step2ab-1}
